A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-10 generalised.tex
comments, fixes

Applied to \cref{const:s1-substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $s_1$ under the substring
operation, and check whether this pre-image is consistent with the constraint~$x
\in \Language_1$, corresponding to the full product seen in
\cref{ex:string-constraints}. Because the substring operation depends on the
values of the integer variables~$i, n$, a Parikh automaton of dimension~$2$,
shown in \cref{fig:aut_a}, models the pre-image. Intuitively, this construction
accumulates any symbol (for the prefix of the substring operation), incrementing
$i$ to mark the start of the substring, followed by the automaton representing
the substring itself, modified to add a counter to increment $n$ at each
transition, followed by a state that performs no increments and only accepts any
suffix after the matched substring.

% @Philipp: I am quite unsure about which intersection we are looking for here,
% please verify I'm not wrong!
Denoting the language described by \cref{fig:aut_a} as $\Language_{pre}$, we can
then replace $s_1 = \Substring(s_2, i, n)$ and $s_2 \in
\Language(B)$ (of \cref{const:s1-in-c-dd}) with an
equi-satisfiable formula that no longer contains any explicit substring
operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language(B)
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    i \\ n
  \end{bmatrix}~~
  \wedge 0 \leq i \leq i + n \leq |p|
\end{equation}
%
To check the satisfiability of \cref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language(B)$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.

>>>>

Applied to \cref{const:s1-substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $s_1$ under the substring
operation, and check whether this pre-image is consistent with the constraint~$s_2 \in \Language(B)$, corresponding to the full product seen in
\cref{ex:string-constraints}. Because the substring operation depends on the
values of the integer variables~$i, n$, a Parikh automaton of dimension~$2$,
shown in \cref{fig:aut_a}, models the pre-image. Intuitively, this construction
accumulates any symbol (for the prefix of the substring operation), incrementing
$i$ to mark the start of the substring, followed by the automaton representing
the substring itself, modified to add a counter to increment $n$ at each
transition, followed by a state that performs no increments and only accepts any
suffix after the matched substring.

% @Philipp: I am quite unsure about which intersection we are looking for here,
% please verify I'm not wrong!
% PR: Sounds right (did some small fixes)
Denoting the language described by \cref{fig:aut_a} as $\Language_{pre}$, we can
then replace $s_1 = \Substring(s_2, i, n)$ and $s_2 \in
\Language(B)$ (of \cref{const:s1-in-c-dd}) with an
equi-satisfiable formula that no longer contains any explicit substring
operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language(B)
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    i \\ n
  \end{bmatrix}~~
  \wedge 0 \leq i \leq i + n \leq |p|
\end{equation}
%
To check the satisfiability of \cref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language(B)$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.
